---
output: html_document
---

```{r setup, include=FALSE}
library(tidyverse)
library(conflicted)

library(pheatmap)

filter <- dplyr::filter

knitr::opts_chunk$set(echo = TRUE)
```

## `pheatmap` Annotations

More detail about heatmap annotations because there were a lot of questions

---

Get the data and drop NAs

```{r}
read_csv('colon.csv') %>%
  na.omit() -> colon
```

Wrangle the data for a heatmap

- drop the categorical columns
- convert the sample IDs or some unique identifier for each row to `rownames`; this isn't 100% necessary, but *if you want to annotate the rows* you need to do this

```{r}
colon %>% 
  select(-id, -rx, -study) %>% 
  as.data.frame() %>%
### need to have rownames (and column names) because that's how pheatmap 
### connects the heatmap table and the annotation table
  column_to_rownames('X1') -> colon_num
```

Plot a heatmap. We can see clustering based on the trees, but we might want to look at some metadata to see if it's overrepresented in some cluster, similar to coloring the points on a PCA plot. That's where annotations come in

```{r}
pheatmap(colon_num, show_rownames = F, scale = 'column')
```

To set up an annotation, you need to set up *a separate annotation table*. The *rownames* in the annotation table need to be the same as the *rownames* in the heatmap table (or if you want to annotate the columns, the *rownames* in your annotation table need to match the *colnames* in the heatmap table)

```{r}
colon %>%
# select the column you used to make the rownames in the heatmap table and any
# columns you want to add as annotations on the heatmap
  select(X1, rx) %>%
  as.data.frame() %>%
# convert the same column you used to make the rownames for the heatmap table
# to the rownames for the annotation table
  column_to_rownames('X1') -> row_anno

# same heatmap code, but with the annotation table added
pheatmap(colon_num, show_rownames = F, scale = 'column', annotation_row = row_anno)
```

You can also add multiple annotations onto the heatmap, although practically speaking more than 2-3 is unreadable

```{r}
colon %>%
# to plot multiple annotations keep multiple columns around
# you can also annotate by continuous variables like age
  select(X1, rx, sex, age) %>%
  mutate(sex = ifelse(sex == 0, 'F', 'M')) %>%
  as.data.frame() %>%
  column_to_rownames('X1') -> row_anno2

pheatmap(colon_num, show_rownames = F, scale = 'column', annotation_row = row_anno2)
```





